\thesistitle

### 2025-07-23 manuscript/main.tex
Acknowl. Contentspage, pagenumbering for pretexts

%=== Metadata ===
\newcommand{\thesistitle}{VS Code Language Extension - MiniProb}
\newcommand{\authorname}{Rafael Tanzer}
\newcommand{\supervisor}{Michele Chiaria, Simone , Francesco Pontigia and Ezzio Bartocci}
\newcommand{\institution}{TU Vienna}
\newcommand{\department}{Cyber-Physical-Systems, ..., ...}
\newcommand{\logoimage}{graphics/logo.png} % Path to your logo image
\newcommand{\submissionmonth}{May}
\newcommand{\submissionyear}{2025}
\newcommand{\submissionday}{24}

@@

%—— German Title Page ——
\begin{titlepage}
  % Top‑left TU Wien logo
  \noindent
  \includegraphics[width=0.25\textwidth]{graphics/tuWienLogo.png}

@@

  % Left‑aligned faculty / advisors block
  \vspace*{1cm}
  \noindent
  an der Fakultät für Informatik\\
  der Technischen Universität Wien\\

@@

%—— English Title Page ——
\begin{titlepage}
  % Top‑left TU Wien logo
  \noindent
  \includegraphics[width=0.25\textwidth]{graphics/tuWienLogo.png}

@@

  % Left‑aligned faculty / advisors block
  \vspace*{1cm}
  \noindent
  to the Faculty of Informatics\\
  at the TU Wien\\

@@

\begin{abstract}
  % Your abstract text here.
  ??all slash code coantined text hs to be revisted due to the detokenize\\

\noindent Cyber-Physical Systems demand rigorous, automated verification to ensure safety and correctness under uncertainty. This thesis presents the design and
implementation of a Visual Studio Code language extension for \textbf{MiniProb} (\ref{sec:miniprob}), a domain-specific probabilistic programming language tailored to formal model checking of
recursive, stochastic processes. Building on the Langium framework - written in TypeScript and running on Node.js - and the Language Server Protocol, a formal EBNF grammar for
MiniProb was defined and used within Langium to generate a parser that produces both Concrete Syntax Trees (CSTs) and typed Abstract Syntax Trees (ASTs).
Custom services - including semantic validation, type checking, cross-reference resolution, and some code completion features - were developed within Langium.

\noindent  The extension, packaged as a .vsix and available on GitHub, streamlines MiniProb development by lowering the barrier to entry for probabilistic model
authors and facilitating direct translation of MiniProb programs into probabilistic operator precedence automata for formal verification.

@@

\end{abstract}

\tableofcontents

@@

conlsuio: mention no dode comülpeteion for structures only refs, not optimal and how it would be implemetned

@@

@@

\begin{bnf}
  $prog$ : \textsf{Program} ::=
  | $decl\_list\; func\; func\_list$
  ;;
  $decl$ : \textsf{Declaration} ::=
  | $type\; identifier\; id\_list$
  | $\epsilon$
  ;;
  $decl\_list$ ::=
  | $decl\; \code{;}\; decl\_list$
  | $\epsilon$
  ;;
  $id\_list$ ::=
  | \code{,}\; $identifier\; id\_list$
  | $\epsilon$
  ;;
  $func$ : \textsf{Function} ::=
  | $identifier\; \left(\; param\_list\; \right)\; \left\{\; decl\_list\; block\; \right\}$
  ;;
  $func\_call$ : \textsf{Func\-Call} ::=
  | $identifier\; \left(\; arg\_list\; \right)$
  ;;
  $arg$ : \textsf{Argument} ::=
  | $expr$
  ;;
  $arg\_list$ ::=
  | $arg\; arg\_list\_tail$
  | $\epsilon$
  ;;
  $arg\_list\_tail$ ::=
  | \code{,} $arg\; arg\_list\_tail$
  | $\epsilon$
  ;;
  $param$ : \textsf{Parameter} ::=
  | $type\;amp\_opt\;identifier$
  ;;
  $amp\_opt$ : \textsf{Value-result} ::=
  | \code{&}
  | $\epsilon$
  ;;
  $param\_list$ ::=
  | $param\;param\_list\_tail$
  | $\epsilon$
  ;;
  $param\_list\_tail$ ::=
  | \code{,} $param\;param\_list\_tail$
  | $\epsilon$
  ;;
  $block$ : \textsf{Codeblock} ::=
  | $stmt\; stmt\_list$
  ;;
  $block\_opt$ ::=
  | $block$
  | $\epsilon$
  ;;
  $stmt$ : \textsf{Statement} ::=
  | $assignment\;$ \code{;}
  | $query\;$ \code{;}
  | $observation\;$ \code{;}
  | $func\_call\;$ \code{;}
  | \code{throw} \code{;}
  | $if\_then\_else$
  | $while$
  | $try\_catch$
  ;;
  $stmt\_list$ ::=
  | $stmt\; stmt\_list$
  | $\epsilon$
  ;;
  $assignment$ ::=
  | $lval\; \code{=}\; assign\_rhs$
  ;;
  $assign\_rhs$  : \textsf{Right-hand side} ::=
  | $expr$
  | $distribution$
  ;;
  $query$ : \textsf{Query} ::=
  | \code{query} $func\_call$
  ;;
  $observation$ : \textsf{Conditioning} ::=
  | \code{observe} $expr$
  ;;
  $if\_then\_else$ : \textsf{IfThenElse} ::=
  | \code{if} $\left(\; expr\; \right) \left\{\; block\_opt\; \right\}$ \code{else} $\left\{\; block\_opt\; \right\}$
  ;;
  $while$ : \textsf{While-loop} ::=
  | \code{while} $\left(\; expr\; \right) \left\{\; block\_opt\; \right\}$
  ;;
  $try\_catch$ : \textsf{Try-catch}::=
  | \code{try} $\left\{\; block\_opt\; \right\} \code{catch} \left\{\; block\_opt\; \right\}$
\end{bnf}

@@

\begin{bnf}
  $identifier$ ::=
  | $letter\;ident\_tail$
  ;;
  $ident\_tail$ ::=
  | $letter\_or\_digit\;ident\_tail$
  | \code{.} $ident\_tail$
  | $\colon\; ident\_tail$
  | \code{~} $ident\_tail$
  | $\epsilon$
  ;;
  $letter$ ::=
  | \code{a}
  | \code{b}
  | …
  | \code{Z}
  | \code{_}
  ;;
  $digit$ ::=
  | \code{0}
  | …
  | \code{9}
  ;;
  $letter\_or\_digit$ ::=
  | $letter$
  | $digit$
  ;;
  $type$ ::=
  | \code{bool}
  | $int\_spec$
  | $int\_spec\;\left[\;int\_literal\;\right]$
  ;;
  $digit\_seq$ ::=
  | $digit\;digit\_seq$
  | $digit$
  ;;
  $int\_spec$ ::=
  | \code{u} $digit\_seq$
  | \code{s} $digit\_seq$
  ;;
  $int\_literal$ ::=
  | $digit\_seq\;int\_spec$
  ;;
  $bool$ ::=
  | \code{true}
  | \code{false}
\end{bnf}
\captionof{listing}{Backus-Naur form of MiniProb grammar}
\label{list:miniprob-bnf}

\chapter{Expanded Code Listings}
\inputminted[linenos, breaklines]{yaml}{samples/mini-prob.langium}
\captionof{listing}{Langium grammar definition of MiniProb (LL(2) parsable)}
\label{list:langium-grammar}

>>>>

%=== Metadata ===
\newcommand{\thesistitle}{VS Code Language Extension - MiniProb}
\newcommand{\authorname}{Rafael Tanzer}
\newcommand{\supervisor}{Michele Chiaria, Simone , Francesco Pontigia and Ezzio Bartocci}
\newcommand{\institution}{TU Vienna}
\newcommand{\department}{Cyber-Physical Systems$_{E191-01}$}
\newcommand{\logoimage}{graphics/logo.png} % Path to your logo image
\newcommand{\submissionmonth}{May}
\newcommand{\submissionyear}{2025}
\newcommand{\submissionday}{24}

@@

%—— German Title Page ——
\begin{titlepage}
  % Top‑left TU Wien logo
  \noindent
  \includegraphics[width=0.3\textwidth]{graphics/tuWienLogo.png}

@@

  % Left‑aligned faculty / advisors block
  \vspace*{1cm}
  \noindent
  an der Fakultät für Informatik: \department\\
  der Technischen Universität Wien\\

@@

  \noindent
  \makebox[\textwidth]{%
    {\small Wien, \submissionday~\submissionmonth~\submissionyear}%
    \hfill
    \begin{tabular}[t]{@{\hspace{2cm}}c@{\hspace{0.2cm}}c@{}}
      \rule{5cm}{0.4pt} & \rule{5cm}{0.4pt} \\[-0.3ex]
      {\small \authorname} & {\small Ezzio Bartocci}
    \end{tabular}%
  }
\end{titlepage}

%—— English Title Page ——
\begin{titlepage}
  % Top‑left TU Wien logo
  \noindent
  \includegraphics[width=0.3\textwidth]{graphics/tuWienLogo.png}

@@

  % Left‑aligned faculty / advisors block
  \vspace*{1cm}
  \noindent
  to the Faculty of Informatics: \department\\
  at the TU Wien\\

@@

%–– Eidesstattliche Erkärung ––%
\thispagestyle{empty}
\begin{flushright}
  {\LARGE\bfseries Erklärung zur Verfassung der Arbeit\par}
\end{flushright}

\vspace{4cm}
\noindent
{\large \authorname}

\vspace{2cm}

% ––– Main declaration text –––
\noindent
Hiermit erkläre ich, dass ich diese Arbeit selbständig verfasst habe, dass ich die verwendeten Quellen und Hilfsmittel vollständig angegeben habe und dass ich die Stellen der Arbeit - einschließlich Tabellen, Karten und Abbildungen -, die anderen Werken oder dem Internet im Wortlaut oder dem Sinn nach entnommen sind, auf jeden Fall unter Angabe der Quelle als Entlehnung kenntlich gemacht habe.

\medskip

\noindent
Ich erkläre weiters, dass ich mich generativer KI-Tools lediglich als Hilfsmittel bedient habe und in der vorliegenden Arbeit mein gestalterischer Einfluss überwiegt. In der "Anerkennung"(Acknowledgements) habe ich alle generativen KI-Tools gelistet, die verwendet wurden, und angegeben, wo und wie sie verwendet wurden. Für Textpassagen, die ohne substantielle Änderungen übernommen wurden, haben ich jeweils die von mir formulierten Eingaben (Prompts) und die verwendete IT-Anwendung mit ihrem Produktnamen und Versionsnummer/Datum angegeben.

\vfill

% ––– Footer: date left, signature right –––
\noindent
\begin{minipage}[t]{0.5\textwidth}
  Wien, \submissionday.~\submissionmonth~\submissionyear
\end{minipage}%
\hfill
\begin{minipage}[t]{0.4\textwidth}
  \hrulefill\\
  \centering \authorname
\end{minipage}

\pagenumbering{roman}
\cleardoublepage
% — Right-aligned, large title —
\phantomsection
\begin{flushright}
  {\Huge\bfseries Abstract\par}
\end{flushright}
\addcontentsline{toc}{chapter}{Abstract}
% Push the block down
\vspace*{3cm}

% Abstract text
\noindent
Cyber-Physical Systems demand rigorous, automated verification to ensure safety and correctness under uncertainty. This thesis presents the design and implementation of
a Visual Studio Code language extension for \textbf{MiniProb} (\ref{sec:miniprob}), a domain-specific probabilistic programming language tailored to formal model
checking of recursive, stochastic processes. Building on the Langium framework - written in TypeScript and running on Node.js - and the Language Server Protocol,
a formal EBNF grammar for MiniProb was defined and used within Langium to generate a parser that produces both Concrete Syntax Trees (CSTs) and
typed Abstract Syntax Trees (ASTs). Custom services - including semantic validation, type checking, cross-reference resolution, and some code completion features -
were developed within Langium.

\vspace{1ex}

\noindent
The extension, packaged as a .vsix and available on \hyperlink{https://github.com/e12224207/miniprob}{GitHub}, streamlines the development of MiniProb programs 
by lowering the barrier to entry for probabilistic model authors and facilitating direct translation of MiniProb programs into probabilistic operator precedence
automata for formal verification.

\vspace{1ex}

@@

\clearpage

\phantomsection
\begin{flushright}
  {\Huge\bfseries \contentsname\par}
\end{flushright}
\addcontentsline{toc}{chapter}{Contents}
\begingroup
  \renewcommand{\contentsname}{\vspace{-1cm}}%
  \tableofcontents
\endgroup

\cleardoublepage
\pagenumbering{arabic}

@@

@@

\clearpage
\chapter{Acknowledgements}

\noindent I would like to thank my supervisory team - Prof.\ First Supervisor, Dr.\ Second Supervisor, and Dr.\ Third Supervisor - for generously sharing their expertise throughout this project.
Their patient guidance on implementation details, contextual foundations, and domain-specific nuances of the application was invaluable. Whenever I encountered a technical
hurdle or needed deeper insight into the domain, they were always ready with thoughtful advice and resources.
\vspace{1ex}
\noindent I would also like to acknowledge the use of ChatGPT 4o-mini-high during the preparation of this thesis. I used it to enhance my writing by correcting grammar
and spelling errors, as well as improve the overall tone and wording.
Furthermore, the AI-tools was used as a consultation on common topics and for creating the Biblatex citations.
All final texts and substantive content, however, reflect my own efforts and have been critically reviewed by me.
\clearpage

@@

\chapter{BNF of MiniProb}
\begin{bnf}
  $prog$ : \textsf{Program} ::=
  | $decl\_list\; func\; func\_list$
  ;;
  $decl$ : \textsf{Declaration} ::=
  | $type\; identifier\; id\_list$
  | $\epsilon$
  ;;
  $decl\_list$ ::=
  | $decl\; \code{;}\; decl\_list$
  | $\epsilon$
  ;;
  $id\_list$ ::=
  | \code{,}\; $identifier\; id\_list$
  | $\epsilon$
  ;;
  $func$ : \textsf{Function} ::=
  | $identifier\; \left(\; param\_list\; \right)\; \left\{\; decl\_list\; block\; \right\}$
  ;;
  $func\_call$ : \textsf{Func\-Call} ::=
  | $identifier\; \left(\; arg\_list\; \right)$
  ;;
  $arg$ : \textsf{Argument} ::=
  | $expr$
  ;;
  $arg\_list$ ::=
  | $arg\; arg\_list\_tail$
  | $\epsilon$
  ;;
  $arg\_list\_tail$ ::=
  | \code{,} $arg\; arg\_list\_tail$
  | $\epsilon$
  ;;
  $param$ : \textsf{Parameter} ::=
  | $type\;amp\_opt\;identifier$
  ;;
  $amp\_opt$ : \textsf{Value-result} ::=
  | \code{&}
  | $\epsilon$
  ;;
  $param\_list$ ::=
  | $param\;param\_list\_tail$
  | $\epsilon$
  ;;
  $param\_list\_tail$ ::=
  | \code{,} $param\;param\_list\_tail$
  | $\epsilon$
  ;;
  $block$ : \textsf{Codeblock} ::=
  | $stmt\; stmt\_list$
  ;;
  $block\_opt$ ::=
  | $block$
  | $\epsilon$
  ;;
  $stmt$ : \textsf{Statement} ::=
  | $assignment\;$ \code{;}
  | $query\;$ \code{;}
  | $observation\;$ \code{;}
  | $func\_call\;$ \code{;}
  | \code{throw} \code{;}
  | $if\_then\_else$
  | $while$
  | $try\_catch$
  ;;
  $stmt\_list$ ::=
  | $stmt\; stmt\_list$
  | $\epsilon$
  ;;
  $assignment$ ::=
  | $lval\; \code{=}\; assign\_rhs$
  ;;
  $assign\_rhs$  : \textsf{Right-hand side} ::=
  | $expr$
  | $distribution$
  ;;
  $query$ : \textsf{Query} ::=
  | \code{query} $func\_call$
  ;;
  $observation$ : \textsf{Conditioning} ::=
  | \code{observe} $expr$
  ;;
  $if\_then\_else$ : \textsf{IfThenElse} ::=
  | \code{if} $\left(\; expr\; \right) \left\{\; block\_opt\; \right\}$ \code{else} $\left\{\; block\_opt\; \right\}$
  ;;
  $while$ : \textsf{While-loop} ::=
  | \code{while} $\left(\; expr\; \right) \left\{\; block\_opt\; \right\}$
  ;;
  $try\_catch$ : \textsf{Try-catch}::=
  | \code{try} $\left\{\; block\_opt\; \right\} \code{catch} \left\{\; block\_opt\; \right\}$
\end{bnf}

@@

\begin{bnf}
  $identifier$ : \textsf{Name} ::=
  | $letter\;ident\_tail$
  ;;
  $ident\_tail$ ::=
  | $letter\_or\_digit\;ident\_tail$
  | \code{.} $ident\_tail$
  | $\colon\; ident\_tail$
  | \code{~} $ident\_tail$
  | $\epsilon$
  ;;
  $letter$ ::=
  | \code{a}
  | \code{b}
  | …
  | \code{Z}
  | \code{_}
  ;;
  $digit$ ::=
  | \code{0}
  | …
  | \code{9}
  ;;
  $letter\_or\_digit$ ::=
  | $letter$
  | $digit$
  ;;
  $type$ : \textsf{Types} ::=
  | \code{bool}
  | $int\_spec$
  | $int\_spec\;\left[\;int\_literal\;\right]$
  ;;
  $digit\_seq$ ::=
  | $digit\;digit\_seq$
  | $digit$
  ;;
  $int\_spec$ : \textsf{Integer} ::=
  | \code{u} $digit\_seq$
  | \code{s} $digit\_seq$
  ;;
  $int\_literal$ ::=
  | $digit\_seq\;int\_spec$
  ;;
  $bool$ : \textsf{Boolean} ::=
  | \code{true}
  | \code{false}
\end{bnf}
\captionof{listing}{Backus-Naur form of MiniProb grammar}
\label{list:miniprob-bnf}

\chapter{section}{Expanded Code Listings}
\inputminted[linenos, breaklines]{yaml}{samples/mini-prob.langium}
\captionof{listing}{Langium grammar definition of MiniProb (LL(2) parsable)}
\label{list:langium-grammar}

### 2025-07-23 manuscript/part1.tex
Acknowl. Contentspage, pagenumbering for pretexts

\subsection{Backus-Naur Form (BNF) and Variants}
Even though production rules, alphabet specifications, and automaton definitions suffice to unambiguously define the set of valid strings/sentences in a language, their notation
tends to be opaque and cumbersome in practice, prompting interest in forming intuitively comprehensible grammar definitions. The Backus-Naur Form (BNF),
created by John Backus with contributions by Peter Naur and released in their Algol-60 report\cite{ALGOL60}, was a pivotal introduction furnishing a concise, human-readable syntax for language generation.
By expressing each rule as
\[
  \langle\mathrm{nonterminal}\rangle \;::=\; expansion_1 \;|\; expansion_2 \;|\;\dots
\]
BNF allows both sequencing and alternation of multiple expansions, enabling language designers to articulate complex, nested structures while preserving structural
clarity and high readability.
This declarative approach not only supports rigorous specification but also facilitates parsing, advancing the practice of compiler development and
language tooling.

>>>>

\subsection{Backus-Naur Form and Variants}
Even though production rules, alphabet specifications, and automaton definitions suffice to unambiguously define the set of valid strings/sentences in a language, their notation
tends to be opaque and cumbersome in practice, prompting interest in forming intuitively comprehensible grammar definitions. The Backus-Naur Form (BNF),
created by John Backus with contributions by Peter Naur and released in their Algol-60 report\cite{ALGOL60}, was a pivotal introduction furnishing a concise, human-readable syntax for language generation.
By expressing each rule as
\[
  \langle\mathrm{nonterminal}\rangle \;::=\; expansion_1 \;|\; expansion_2 \;|\;\dots
\]
BNF allows both sequencing and alternation of multiple expansions, enabling language designers to articulate complex, nested structures while preserving structural
clarity and high readability.
This declarative approach not only supports rigorous specification but also facilitates parsing, advancing the practice of compiler development and
language tooling.
